Avoiding tipping points in fisheries management through Gaussian Process Dynamic Programming

### 2014-10-15 manuscripts/manuscript.Rmd
rerun

<!-- 
### Figure 1: Fitted Models

- All models fit the data quite well
- Information criteria would pick the simple, incorrect model.
-->

@@

We show the  mean inferred state space dynamics of each model relative to the
true model used to generate the data in Figure 2, predicting
the relationship between observed stock size (x-axis) to the stock
size after recruitment the following year.  Note that in contrast to
the  other models shown, the expected Gaussian process corresponds to a
distribution of curves - as indicated by the gray band - which itself
has a mean shown in black. Parameter uncertainty (not shown) spreads
out the estimates further.  The observed data from which each model are
estimated is also shown.  The observations come from only a limited region
of state space corresponding to unharvested or weakly harvested system.
No observations occur at the theoretical optimum harvest rate or near
the tipping point.

Figure 3 illustrates the performance of the GP and parametric models
outside the observed training data. The mean trajectory under the
underlying model is shown by the black dots, while the corresponding
prediction made by the model shown by the box and whiskers plots.
Predictions are based on the true expected value in the previous time
step.  Predicted distributions that lie entirely above the expected
dynamics indicate the expectation of stock sizes higher than what is
actually expected. The models differ both in their expectations and
their uncertainty (colored bands show two standard deviations away).
Note that the GP is particularly uncertain about the dynamics relative
to structurally incorrect models like the Ricker.

```{r figure_3, fig.width=8, fig.height=6, fig.cap="Outside the range of the training data (Figure 1), the true dynamics (black dots) fall outside the uncertainty (two standard deviations, colored bands) of the structurally incorrect parametric models (Ricker, Myers), but inside the uncertainty predicted by the GP. Points show the stock size simulated by the true model.  Overlay shows the range of states predicted by each model, based on the state observed in the previous time step. The Ricker model always (wrongly) predicts positive population growth, while the actual population shrinks in each step as the initial condition falls below the Allee threshold of the underlying model (Allen).  Note that because it does not assume a parametric form but instead relies more directly on the data, the GP is both more pessimistic and more uncertain about the future state than the parametric models.", dependson="plot-options"}
y <- numeric(8)
y[1] <- 4.5
for(t in 1:(length(y)-1))
      y[t+1] = z_g() * f(y[t], h=0, p=p)
# predicts means, does not reflect uncertainty estimate!
crash_data <- step_ahead_posteriors(y)
crash_data <- subset(crash_data, variable %in% c("GP", "Allen", "Ricker", "Myers"))
ggplot(crash_data) + 
  geom_boxplot(aes(as.factor(as.integer(time)), value, 
                   fill = variable, col=variable), 
               outlier.size=1, position="identity") + 
#  geom_line(aes(time, value, col = variable, 
#            group=interaction(L1,variable))) + 
  geom_point(aes(time, stock), size = 3) + 
  scale_fill_manual(values=colorkey[c("GP", "Allen", "Ricker", "Myers")]) +  
  scale_colour_manual(values=colorkey[c("GP", "Allen", "Ricker", "Myers")]) +  
  facet_wrap(~variable) + 
  theme(legend.position="none") + xlab("time") + ylab("stock size")

write.csv(crash_data, "components/data/figure3.csv")
```

<!--
### Figure 4: Inferred Policies

- Inferred policies differ substantially among models
- The structurally correct model and the GP are close to the true model
- alternatives are not close
-->

@@

Despite the similarities in model fits to the observed data, the
policies inferred under each model differ widely (Figure 4).
Policies are shown in terms of target escapement, $S_t$.  Under models
such as this a constant escapement policy is expected to be optimal
[@Reed1979], whereby population levels below a certain size $S$ are
unharvested, while above that size the harvest strategy aims to return the
population to $S$, resulting in the hockey-stick shaped policies shown.
Only the structurally correct model (Allen model) and the GP produce
policies close to the true optimum policy (where both the underlying
model structure and parameter values are known without error).

@@

In Figure 5, we show the consequences of managing 100 replicate
realizations of the simulated fishery under each of the policies
estimated.  As anticipated from the policy curves, the structurally
correct model under-harvests, leaving the stock to vary around its
unfished optimum.  The structurally incorrect Ricker model over-harvests
the population passed the tipping point consistently, resulting in the
immediate crash of the stock and thus derives minimal long-term catch.

The results across replicate stochastic simulations can most
easily be compared by using the relative differences in net present
value realized by each of the model (Figure 6).  The GPDP
most consistently realizes a value close to the optimal solution,
and importantly avoids ever driving the system across the tipping point,
which results in the near-zero value cases in the parametric models.

@@

These results are not sensitive to the modeling details of the simulation.
The GPDP estimate remains very close to the optimal solution obtained
by knowing the true model across changes to the training simulation,
noise scale, parameters or structure of the underlying model. We consider
both a Latin hypercube approach and a more focused investigation of the
parameters most responsible for impacting the transition rate: the relative 
distance to the Allee threshold and the noise effects.

The Latin hypercube approach systematically varies all combinations of
parameters, providing a more general test than varying only one parameter
at a time.  We loop across eight replicates of three different randomly
generated parameter sets for each of two different generating models
(Allen and Myers) over two different noise levels (0.01 and 0.05),
for a total of 8 x 3 x 2 x 2 = 96 scenarios. The Gaussian Process performs
nearly optimally in each case, relative to the optimal solution with no
parameter or model uncertainty (Figure S9, appendix).

@@

The GPDP is weakly influenced by increasing stochasticity or increasing
Allee effects over much of the range (Figure 7) Much larger stochasticity or higher Allee
effects make even the optimal solution without any model or parameter uncertainty
unable to harvest the population effectively (e.g. noise large enough to violate the
self-sustaining criterion of @Reed1979). Only at the largest Allee effect
sizes are some replicates seen to deviate more from the optimal solution.

>>>>

@@

We show the  mean inferred state space dynamics of each model relative to the
true model used to generate the data in Figure 2, predicting
the relationship between observed stock size (x-axis) to the stock
size after recruitment the following year.  Note that in contrast to
the  other models shown, the expected Gaussian process corresponds to a
distribution of curves - as indicated by the gray band - which itself
has a mean shown in black. Parameter uncertainty (not shown) spreads
out the estimates further.  The observed data from which each model are
estimated is also shown.  The observations come from only a limited region
of state space corresponding to unharvested or weakly harvested system.
No observations occur at the theoretical optimum harvest rate or near
the tipping point. Figure S1 (see supplement) shows the performance of the models outside the observed training data.

@@

Despite the similarities in model fits to the observed data, the
policies inferred under each model differ widely (Figure 3).
Policies are shown in terms of target escapement, $S_t$.  Under models
such as this a constant escapement policy is expected to be optimal
[@Reed1979], whereby population levels below a certain size $S$ are
unharvested, while above that size the harvest strategy aims to return the
population to $S$, resulting in the hockey-stick shaped policies shown.
Only the structurally correct model (Allen model) and the GP produce
policies close to the true optimum policy (where both the underlying
model structure and parameter values are known without error).

@@

In Figure 4, we show the consequences of managing 100 replicate
realizations of the simulated fishery under each of the policies
estimated.  As anticipated from the policy curves, the structurally
correct model under-harvests, leaving the stock to vary around its
unfished optimum.  The structurally incorrect Ricker model over-harvests
the population passed the tipping point consistently, resulting in the
immediate crash of the stock and thus derives minimal long-term catch.

The results across replicate stochastic simulations can most
easily be compared by using the relative differences in net present
value realized by each of the model (Figure 5).  The GPDP
most consistently realizes a value close to the optimal solution,
and importantly avoids ever driving the system across the tipping point,
which results in the near-zero value cases in the parametric models.

@@

These results are not sensitive to the modeling details of the simulation.
The GPDP estimate remains very close to the optimal solution obtained
by knowing the true model across changes to the training simulation,
noise scale, parameters or structure of the underlying model. We consider
both a Latin hypercube approach (see supplementary materials, Fig S11) and a more focused investigation of the
parameters most responsible for impacting the transition rate: the relative 
distance to the Allee threshold and the noise effects.

@@

The GPDP is only weakly influenced by increasing stochasticity or increasing
Allee effects over much of the range (Figure 6). Much larger stochasticity or higher Allee
effects make even the optimal solution without any model or parameter uncertainty
unable to harvest the population effectively (e.g. noise large enough to violate the
self-sustaining criterion of @Reed1979). Only at the largest Allee effect
sizes are some replicates seen to deviate more from the optimal solution.

### 2014-10-15 manuscripts/manuscript.tex
rerun

We show the mean inferred state space dynamics of each model relative to
the true model used to generate the data in Figure 2, predicting the
relationship between observed stock size (x-axis) to the stock size
after recruitment the following year. Note that in contrast to the other
models shown, the expected Gaussian process corresponds to a
distribution of curves - as indicated by the gray band - which itself
has a mean shown in black. Parameter uncertainty (not shown) spreads out
the estimates further. The observed data from which each model are
estimated is also shown. The observations come from only a limited
region of state space corresponding to unharvested or weakly harvested
system. No observations occur at the theoretical optimum harvest rate or
near the tipping point.

Figure 3 illustrates the performance of the GP and parametric models
outside the observed training data. The mean trajectory under the
underlying model is shown by the black dots, while the corresponding
prediction made by the model shown by the box and whiskers plots.
Predictions are based on the true expected value in the previous time
step. Predicted distributions that lie entirely above the expected
dynamics indicate the expectation of stock sizes higher than what is
actually expected. The models differ both in their expectations and
their uncertainty (colored bands show two standard deviations away).
Note that the GP is particularly uncertain about the dynamics relative
to structurally incorrect models like the Ricker.

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_3.pdf}
\caption{Outside the range of the training data (Figure 1), the true
dynamics (black dots) fall outside the uncertainty (two standard
deviations, colored bands) of the structurally incorrect parametric
models (Ricker, Myers), but inside the uncertainty predicted by the GP.
Points show the stock size simulated by the true model. Overlay shows
the range of states predicted by each model, based on the state observed
in the previous time step. The Ricker model always (wrongly) predicts
positive population growth, while the actual population shrinks in each
step as the initial condition falls below the Allee threshold of the
underlying model (Allen). Note that because it does not assume a
parametric form but instead relies more directly on the data, the GP is
both more pessimistic and more uncertain about the future state than the
parametric models.}
\end{figure}

@@

Despite the similarities in model fits to the observed data, the
policies inferred under each model differ widely (Figure 4). Policies
are shown in terms of target escapement, $S_t$. Under models such as
this a constant escapement policy is expected to be optimal (Reed 1979),
whereby population levels below a certain size $S$ are unharvested,
while above that size the harvest strategy aims to return the population
to $S$, resulting in the hockey-stick shaped policies shown. Only the
structurally correct model (Allen model) and the GP produce policies
close to the true optimum policy (where both the underlying model
structure and parameter values are known without error).

@@

The GPDP is weakly influenced by increasing stochasticity or increasing
Allee effects over much of the range (Figure 7) Much larger
stochasticity or higher Allee effects make even the optimal solution
without any model or parameter uncertainty unable to harvest the
population effectively (e.g.~noise large enough to violate the
self-sustaining criterion of Reed (1979)). Only at the largest Allee
effect sizes are some replicates seen to deviate more from the optimal
solution.

>>>>

We show the mean inferred state space dynamics of each model relative to
the true model used to generate the data in Figure 2, predicting the
relationship between observed stock size (x-axis) to the stock size
after recruitment the following year. Note that in contrast to the other
models shown, the expected Gaussian process corresponds to a
distribution of curves - as indicated by the gray band - which itself
has a mean shown in black. Parameter uncertainty (not shown) spreads out
the estimates further. The observed data from which each model are
estimated is also shown. The observations come from only a limited
region of state space corresponding to unharvested or weakly harvested
system. No observations occur at the theoretical optimum harvest rate or
near the tipping point. Figure S1 (see supplement) shows the performance
of the models outside the observed training data.

@@

Despite the similarities in model fits to the observed data, the
policies inferred under each model differ widely (Figure 3). Policies
are shown in terms of target escapement, $S_t$. Under models such as
this a constant escapement policy is expected to be optimal (Reed 1979),
whereby population levels below a certain size $S$ are unharvested,
while above that size the harvest strategy aims to return the population
to $S$, resulting in the hockey-stick shaped policies shown. Only the
structurally correct model (Allen model) and the GP produce policies
close to the true optimum policy (where both the underlying model
structure and parameter values are known without error).

@@

The GPDP is only weakly influenced by increasing stochasticity or
increasing Allee effects over much of the range (Figure 6). Much larger
stochasticity or higher Allee effects make even the optimal solution
without any model or parameter uncertainty unable to harvest the
population effectively (e.g.~noise large enough to violate the
self-sustaining criterion of Reed (1979)). Only at the largest Allee
effect sizes are some replicates seen to deviate more from the optimal
solution.
